---
title: 3/5740 Final Project
output: html_document
---

Group Members:

- Alisa Castillo (alc297)
- JP Eliopoulos (jae257)
- Cole Garcia (cgg38)
- Allyanna Panganiban (amp388)
- Sarah Swee (sjs469)


## Data Pre-processing
```{r load packages}
# be sure to have installed the necessary packages prior to running this cell 
# (see `README.md`)
library(knitr)
library(rmarkdown)
library(tidyverse)
library(dplyr)
library(readr)
library(caret)
library(pROC)
library(broom)
```


```{r load data}
diabetic = read.csv("data/diabetic_data.csv", na=c("?","NA"))
head(diabetic)
mapping = read.csv("data/IDS_mapping.csv")
head(mapping)
```


```{r pre-processing from paper}
# remove high-missing & irrelevant columns
diabetic_clean <- diabetic %>%
  select(-any_of(c("weight", "payer_code"))) %>%
  mutate(
    medical_specialty = ifelse(is.na(medical_specialty), "Missing", medical_specialty),
    race = ifelse(is.na(race), "Missing", race)
  )
head(diabetic_clean)

# keep only first encounter per patient
diabetic_clean <- diabetic_clean %>%
  arrange(patient_nbr, encounter_id) %>%
  distinct(patient_nbr, .keep_all = TRUE)
cat("after keeping first encounter per patient:", dim(diabetic_clean), "\n")

# remove death/hospice discharges
death_hospice_ids <- c(11, 13, 14, 19, 20, 21)
diabetic_clean <- diabetic_clean %>%
  filter(!discharge_disposition_id %in% death_hospice_ids)
cat("after removing death/hospice:", dim(diabetic_clean), "\n")

# check whether an ICD-9 diagnosis code indicates diabetes
is_diabetes_dx <- function(code) {
  code <- as.character(code)
  ifelse(
    is.na(code),
    FALSE,
    startsWith(code,"250")
  )
}

diabetic_clean <- diabetic_clean %>%
  filter(
    # paper keeps stays between 1 and 14 days
    time_in_hospital >= 1,
    time_in_hospital <= 14,
    # paper requires lab tests during the encounter
    num_lab_procedures > 0,
    # paper requires medications during the encounter
    num_medications > 0,
    # paper keeps diabetic encounters:
    # diabetes appears in any diagnosis field
    is_diabetes_dx(diag_1) |
      is_diabetes_dx(diag_2) |
      is_diabetes_dx(diag_3)
  )
```

```{r pre-processing from paper2}
# creating binary readmission outcome
diabetic_clean <- diabetic_clean %>%
  mutate(
    readmitted_binary = ifelse(readmitted == "<30",1,0)
  )

# create HbA1c group
# HbA1c = a blood test measuring average blood sugar over ~3 months
# paper has:
# no HbA1c test
# normal result
# high result and medication changed
# high result and medication not changed
diabetic_clean <- diabetic_clean %>%
  mutate(
    hba1c_group = case_when(
      A1Cresult == "None"~"No test",
      A1Cresult == "Norm"~"Normal",
      A1Cresult %in% c(">7",">8") & change == "Ch"~"High, medication changed",
      A1Cresult %in% c(">7",">8") & change == "No"~"High, medication not changed",
      TRUE ~ NA_character_
    )
  )
head(diabetic_clean)
```

```{r pre-processing from paper3: re-coding}
# recoding age into age groups specified in paper
diabetic_clean <- diabetic_clean %>%
  mutate(
    age_group_paper = case_when(
      age %in% c("[0-10)", "[10-20)", "[20-30)") ~ "<=30",
      age %in% c("[30-40)", "[40-50)", "[50-60)") ~ "30-60",
      age %in% c("[60-70)", "[70-80)", "[80-90)", "[90-100)") ~ ">60",
      TRUE ~ NA_character_
    )
  )

# recoding primary diagnosis (encoded) into 8 meaningful groups
diag_category <- function(code) {
  if (is.na(code)) {
    return("Other")
  }
  code <- as.character(code)
  if (startsWith(code, "V") || startsWith(code, "E")) {
    return("Other")
  }
  num <- suppressWarnings(as.numeric(code))
  if (is.na(num)) {
    return("Other")
  } else if ((num>=390 & num<=459) | num==785) {
    return("Circulatory")
  } else if ((num>=460 & num<=519) | num==786) {
    return("Respiratory")
  } else if ((num>=520 & num<=579) | num==787) {
    return("Digestive")
  } else if (floor(num)==250) {
    return("Diabetes")
  } else if (num>=800 & num<=999) {
    return("Injury")
  } else if (num>=710 & num<=739) {
    return("Musculoskeletal")
  } else if ((num>=580 & num<=629) | num==788) {
    return("Genitourinary")
  } else if (num>=140 & num<=239) {
    return("Neoplasms")
  } else {
    return("Other")
  }
}

diabetic_clean <- diabetic_clean %>%
  mutate(
    primary_diagnosis_group = sapply(diag_1, diag_category)
  )
head(diabetic_clean)

# recoding admission source
diabetic_clean <- diabetic_clean %>%
  mutate(
    admission_source_group = case_when(
      admission_source_id == 7 ~ "Emergency room",
      admission_source_id == 1 ~ "Physician/clinic referral",
      TRUE ~ "Otherwise"
    )
  )

# recoding discharge disposition
diabetic_clean <- diabetic_clean %>%
  mutate(
    discharge_home = ifelse(
      discharge_disposition_id==1,
      "Discharged to home",
      "Otherwise"
    )
  )

# recoding medical specialty into main groups
surgery_specialties <- c(
  "Surgery-General",
  "Surgery-Cardiovascular/Thoracic",
  "Surgery-Neuro",
  "Surgery-Vascular",
  "Surgery-Plastic",
  "Surgery-Cardiovascular",
  "Surgery-Thoracic",
  "Surgery-Colon&Rectal",
  "Surgery-Pediatric",
  "Surgery-Maxillofacial"
)
diabetic_clean <- diabetic_clean %>%
  mutate(
    medical_specialty_group = case_when(
      is.na(medical_specialty) | medical_specialty == "Missing" ~ "Missing or unknown",
      medical_specialty == "InternalMedicine" ~ "Internal Medicine",
      medical_specialty == "Cardiology" ~ "Cardiology",
      medical_specialty %in% surgery_specialties ~ "Surgery",
      medical_specialty == "Family/GeneralPractice" ~ "Family/general practice",
      TRUE ~ "Other"
    )
  )
```

```{r pre-processing final}
# selecting final variables
final_cols <- c(
  "encounter_id",
  "patient_nbr",
  "readmitted",
  "readmitted_binary",
  "hba1c_group",
  "gender",
  "age",
  "age_group_paper",
  "race",
  "admission_source_id",
  "admission_source_group",
  "discharge_disposition_id",
  "discharge_home",
  "diag_1",
  "primary_diagnosis_group",
  "medical_specialty",
  "medical_specialty_group",
  "time_in_hospital"
)

diabetic_clean <- diabetic_clean %>%
  select(any_of(final_cols))

cat("final shape:", dim(diabetic_clean), "\n")
head(diabetic_clean)
```

## Modeling

### A. Logistic Regression

```{r logistic}
# preparing data for logistic -----------------------------------------------
model_data <- diabetic_clean %>%
  select(
    readmitted_binary,
    hba1c_group,
    age_group_paper,
    race,
    admission_source_group,
    discharge_home,
    primary_diagnosis_group,
    medical_specialty_group,
    time_in_hospital
  ) %>%
  mutate(
    readmitted_binary = factor(
      readmitted_binary,
      levels = c(0, 1),
      labels = c("Not_Readmitted", "Readmitted")
    ),
    across(
      c(hba1c_group, age_group_paper, race, admission_source_group,
        discharge_home, primary_diagnosis_group, medical_specialty_group),
      as.factor
    )
  )

set.seed(123)

train_index <- createDataPartition(
  model_data$readmitted_binary,
  p = 0.70,
  list = FALSE
)

train_data <- model_data[train_index, ]
test_data  <- model_data[-train_index, ]

# logistic regression model ------------------------------------------------
logistic <- glm(
  readmitted_binary ~ discharge_home +
    race +
    admission_source_group +
    medical_specialty_group +
    time_in_hospital +
    age_group_paper +
    primary_diagnosis_group,
  data = train_data,
  family = binomial
)
summary(logistic)

# add HbA1c group
hba1c_logistic <- glm(
  readmitted_binary ~ discharge_home +
    race +
    admission_source_group +
    medical_specialty_group +
    time_in_hospital +
    age_group_paper +
    primary_diagnosis_group +
    hba1c_group,
  data = train_data,
  family = binomial
)
summary(hba1c_logistic)

# compare core model vs HbA1c logistic ------------------------------------
anova(logistic, hba1c_logistic, test = "Chisq")

# interaction terms -------------------------------------------------------
interaction_model <- glm(
  readmitted_binary ~ discharge_home +
    race +
    admission_source_group +
    medical_specialty_group +
    time_in_hospital +
    age_group_paper +
    primary_diagnosis_group +
    hba1c_group +
    # interactions described in the paper
    discharge_home:race +
    discharge_home:medical_specialty_group +
    discharge_home:primary_diagnosis_group +
    discharge_home:time_in_hospital +
    medical_specialty_group:time_in_hospital +
    medical_specialty_group:age_group_paper +
    primary_diagnosis_group:time_in_hospital +
    primary_diagnosis_group:hba1c_group,
  data = train_data,
  family = binomial
)
summary(interaction_model)

# Compare HbA1c model vs interaction model
anova(hba1c_logistic, interaction_model, test = "Chisq")

# prediction on test set ----------------------------------------------------
test_data$prob_readmit <- predict(
  interaction_model,
  newdata = test_data,
  type = "response"
)

# ROC / AUC ----------------------------------------------------------------
roc_obj <- roc(
  response = test_data$readmitted_binary,
  predictor = test_data$prob_readmit,
  levels = c("Not_Readmitted", "Readmitted")
)

auc(roc_obj)

# choose threshold to prioritize recall/sensitivity, 70% sensitivity -------------------------
all_thresholds <- coords(
  roc_obj,
  x = "all",
  ret = c("threshold", "sensitivity", "specificity"),
  transpose = FALSE
)
all_thresholds <- all_thresholds %>%
  filter(!is.na(threshold)) %>%
  mutate(diff_from_target = abs(sensitivity - 0.70)) %>%
  arrange(diff_from_target)

threshold_info <- all_thresholds[1, ]
threshold_info
threshold <- threshold_info$threshold
threshold

# classify using recall-focused threshold ----------------------------------
test_data$pred_class <- ifelse(
  test_data$prob_readmit >= threshold,
  "Readmitted",
  "Not_Readmitted"
)

test_data$pred_class <- factor(
  test_data$pred_class,
  levels = c("Not_Readmitted", "Readmitted")
)
```

```{r logistic_model evaluation}
# model evaluation -----------------------------------------
cm <- confusionMatrix(
  test_data$pred_class,
  test_data$readmitted_binary,
  positive = "Readmitted"
)
cm
auc(roc_obj)
plot(roc_obj, main = "ROC Curve: Logistic Regression")
abline(a = 0, b = 1, lty = 2)
```

The main logistic regression model was extended by adding HbA1c and interaction 
terms. These different models were then compared to each other. A likelihood ratio 
test  showed that HbA1c did not significantly improve  model fit (p=0.32). However, 
the interaction model significantly reduced deviance compared to the HbA1c 
model (p<0.001), suggesting improved fit.

Due to class imbalance (only ~8.6% of patients were readmitted), using a default
classification threshold of 0.5 resulted in a model that failed to identify 
ny readmissions. To address this, we selected a lower threshold (~0.072) 
based on the ROC curve to achieve approximately 70% sensitivity.

At this threshold, the cost of reduced specificity (~47%) and precision (~11%), 
reflected an increase in false positives. However, in a healthcare context, 
false negatives (missed readmissions) are more costly than false positives, 
making may deem this tradeoff appropriate. The model achieved an AUC of 0.623, 
indicating a modest predictive performance.

