---
title: 3/5740 Final Project
output: html_document
---

Group Members:

- Alisa Castillo (alc297)
- JP Eliopoulos (jae257)
- Cole Garcia (cgg38)
- Allyanna Panganiban (amp388)
- Sarah Swee (sjs469)


```{r load packages}
# be sure to have installed the necessary packages prior to running this cell 
# (see `README.md`)
# install.packages("class", "caret")
library(knitr)
library(rmarkdown)
library(tidyverse)
library(dplyr)
library(readr)
library(glmnet)
library(caret)
library(pROC)
library(gam)
library(class)   
```


```{r load data}
diabetic = read.csv("data/diabetic_data.csv", na=c("?","NA"))
head(diabetic)
mapping = read.csv("data/IDS_mapping.csv")
head(mapping)
```


```{r pre-processing from paper}
# remove high-missing & irrelevant columns
diabetic_clean <- diabetic %>%
  select(-any_of(c("weight", "payer_code"))) %>%
  mutate(
    medical_specialty = ifelse(is.na(medical_specialty), "Missing", medical_specialty),
    race = ifelse(is.na(race), "Missing", race)
  )
head(diabetic_clean)

# keep only first encounter per patient
diabetic_clean <- diabetic_clean %>%
  arrange(patient_nbr, encounter_id) %>%
  distinct(patient_nbr, .keep_all = TRUE)
cat("after keeping first encounter per patient:", dim(diabetic_clean), "\n")

# remove death/hospice discharges
death_hospice_ids <- c(11, 13, 14, 19, 20, 21)
diabetic_clean <- diabetic_clean %>%
  filter(!discharge_disposition_id %in% death_hospice_ids)
cat("after removing death/hospice:", dim(diabetic_clean), "\n")

# check whether an ICD-9 diagnosis code indicates diabetes
is_diabetes_dx <- function(code) {
  code <- as.character(code)
  ifelse(
    is.na(code),
    FALSE,
    startsWith(code,"250")
  )
}

diabetic_clean <- diabetic_clean %>%
  filter(
    # paper keeps stays between 1 and 14 days
    time_in_hospital >= 1,
    time_in_hospital <= 14,
    # paper requires lab tests during the encounter
    num_lab_procedures > 0,
    # paper requires medications during the encounter
    num_medications > 0,
    # paper keeps diabetic encounters:
    # diabetes appears in any diagnosis field
    is_diabetes_dx(diag_1) |
      is_diabetes_dx(diag_2) |
      is_diabetes_dx(diag_3)
  )
```

```{r pre-processing from paper2}
# creating binary readmission outcome
diabetic_clean <- diabetic_clean %>%
  mutate(
    readmitted_binary = ifelse(readmitted == "<30",1,0)
  )

# create HbA1c group
# HbA1c = a blood test measuring average blood sugar over ~3 months
# paper has:
# no HbA1c test
# normal result
# high result and medication changed
# high result and medication not changed
diabetic_clean <- diabetic_clean %>%
  mutate(
    hba1c_group = case_when(
      A1Cresult == "None"~"No test",
      A1Cresult == "Norm"~"Normal",
      A1Cresult %in% c(">7",">8") & change == "Ch"~"High, medication changed",
      A1Cresult %in% c(">7",">8") & change == "No"~"High, medication not changed",
      TRUE ~ NA_character_
    )
  )
head(diabetic_clean)
```

```{r pre-processing from paper3: re-coding}
# recoding age into age groups specified in paper
diabetic_clean <- diabetic_clean %>%
  mutate(
    age_group_paper = case_when(
      age %in% c("[0-10)", "[10-20)", "[20-30)") ~ "<=30",
      age %in% c("[30-40)", "[40-50)", "[50-60)") ~ "30-60",
      age %in% c("[60-70)", "[70-80)", "[80-90)", "[90-100)") ~ ">60",
      TRUE ~ NA_character_
    )
  )

# recoding primary diagnosis (encoded) into 8 meaningful groups
diag_category <- function(code) {
  if (is.na(code)) {
    return("Other")
  }
  code <- as.character(code)
  if (startsWith(code, "V") || startsWith(code, "E")) {
    return("Other")
  }
  num <- suppressWarnings(as.numeric(code))
  if (is.na(num)) {
    return("Other")
  } else if ((num>=390 & num<=459) | num==785) {
    return("Circulatory")
  } else if ((num>=460 & num<=519) | num==786) {
    return("Respiratory")
  } else if ((num>=520 & num<=579) | num==787) {
    return("Digestive")
  } else if (floor(num)==250) {
    return("Diabetes")
  } else if (num>=800 & num<=999) {
    return("Injury")
  } else if (num>=710 & num<=739) {
    return("Musculoskeletal")
  } else if ((num>=580 & num<=629) | num==788) {
    return("Genitourinary")
  } else if (num>=140 & num<=239) {
    return("Neoplasms")
  } else {
    return("Other")
  }
}

diabetic_clean <- diabetic_clean %>%
  mutate(
    primary_diagnosis_group = sapply(diag_1, diag_category)
  )
head(diabetic_clean)

# recoding admission source
diabetic_clean <- diabetic_clean %>%
  mutate(
    admission_source_group = case_when(
      admission_source_id == 7 ~ "Emergency room",
      admission_source_id == 1 ~ "Physician/clinic referral",
      TRUE ~ "Otherwise"
    )
  )

# recoding discharge disposition
diabetic_clean <- diabetic_clean %>%
  mutate(
    discharge_home = ifelse(
      discharge_disposition_id==1,
      "Discharged to home",
      "Otherwise"
    )
  )

# recoding medical specialty into main groups
surgery_specialties <- c(
  "Surgery-General",
  "Surgery-Cardiovascular/Thoracic",
  "Surgery-Neuro",
  "Surgery-Vascular",
  "Surgery-Plastic",
  "Surgery-Cardiovascular",
  "Surgery-Thoracic",
  "Surgery-Colon&Rectal",
  "Surgery-Pediatric",
  "Surgery-Maxillofacial"
)
diabetic_clean <- diabetic_clean %>%
  mutate(
    medical_specialty_group = case_when(
      is.na(medical_specialty) | medical_specialty == "Missing" ~ "Missing or unknown",
      medical_specialty == "InternalMedicine" ~ "Internal Medicine",
      medical_specialty == "Cardiology" ~ "Cardiology",
      medical_specialty %in% surgery_specialties ~ "Surgery",
      medical_specialty == "Family/GeneralPractice" ~ "Family/general practice",
      TRUE ~ "Other"
    )
  )
```

```{r pre-processing final}
# selecting final variables
final_cols <- c(
  "encounter_id",
  "patient_nbr",
  "readmitted",
  "readmitted_binary",
  "hba1c_group",
  "gender",
  "age",
  "age_group_paper",
  "race",
  "admission_source_id",
  "admission_source_group",
  "discharge_disposition_id",
  "discharge_home",
  "diag_1",
  "primary_diagnosis_group",
  "medical_specialty",
  "medical_specialty_group",
  "time_in_hospital"
)

diabetic_clean <- diabetic_clean %>%
  select(any_of(final_cols))

cat("final shape:", dim(diabetic_clean), "\n")
head(diabetic_clean)
```

## LASSO/Ridge Regression

```{r}
# Use only cleaned/grouped predictors
model_data <- diabetic_clean |>
  select(
    readmitted_binary,
    hba1c_group,
    gender,
    age_group_paper,
    race,
    admission_source_group,
    discharge_home,
    primary_diagnosis_group,
    medical_specialty_group,
    time_in_hospital
  ) |>
  drop_na()

# Make outcome numeric
model_data$readmitted_binary <- as.numeric(model_data$readmitted_binary)
```

```{r}
set.seed(1)

train_index <- createDataPartition(
  model_data$readmitted_binary,
  p = 0.7,
  list = FALSE
)

train_data <- model_data[train_index, ]
test_data  <- model_data[-train_index, ]
```

```{r}
# glmnet needs predictors as a matrix
x_train <- model.matrix(readmitted_binary ~ ., data = train_data)[, -1]
y_train <- train_data$readmitted_binary

x_test <- model.matrix(readmitted_binary ~ ., data = test_data)[, -1]
y_test <- test_data$readmitted_binary
```

```{r}
set.seed(1)

ridge_model <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 0,
  family = "binomial"
)

best_lambda_ridge <- ridge_model$lambda.min
best_lambda_ridge

ridge_probs <- predict(
  ridge_model,
  s = best_lambda_ridge,
  newx = x_test,
  type = "response"
)

ridge_class <- ifelse(ridge_probs > 0.5, 1, 0)
```

```{r}
set.seed(1)

lasso_model <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 1,
  family = "binomial"
)

best_lambda_lasso <- lasso_model$lambda.min
best_lambda_lasso

lasso_probs <- predict(
  lasso_model,
  s = best_lambda_lasso,
  newx = x_test,
  type = "response"
)

lasso_class <- ifelse(lasso_probs > 0.5, 1, 0)
```

```{r}
# Confusion matrices
ridge_cm <- confusionMatrix(
  factor(ridge_class, levels = c(0, 1)),
  factor(y_test, levels = c(0, 1)),
  positive = "1"
)

lasso_cm <- confusionMatrix(
  factor(lasso_class, levels = c(0, 1)),
  factor(y_test, levels = c(0, 1)),
  positive = "1"
)

ridge_cm
lasso_cm
```

```{r}
# AUC values
ridge_auc <- auc(y_test, as.numeric(ridge_probs))
lasso_auc <- auc(y_test, as.numeric(lasso_probs))

ridge_auc
lasso_auc
```

```{r}
comparison_table <- tibble(
  Model = c("Ridge Regression", "LASSO Regression"),
  Best_Lambda = c(best_lambda_ridge, best_lambda_lasso),
  Accuracy = c(
    ridge_cm$overall["Accuracy"],
    lasso_cm$overall["Accuracy"]
  ),
  Sensitivity = c(
    ridge_cm$byClass["Sensitivity"],
    lasso_cm$byClass["Sensitivity"]
  ),
  Specificity = c(
    ridge_cm$byClass["Specificity"],
    lasso_cm$byClass["Specificity"]
  ),
  AUC = c(ridge_auc, lasso_auc)
)

comparison_table
```


```{r}
lasso_coef <- coef(lasso_model, s = best_lambda_lasso)

lasso_coef_df <- as.data.frame(as.matrix(lasso_coef)) %>%
  rownames_to_column("Variable") %>%
  rename(Coefficient = 2) %>%
  filter(Coefficient != 0)

lasso_coef_df
```

```{r}
ridge_coef <- coef(ridge_model, s = best_lambda_ridge)

ridge_coef_df <- as.data.frame(as.matrix(ridge_coef)) %>%
  rownames_to_column("Variable") %>%
  rename(Coefficient = 2)

ridge_coef_df
```

Ridge and LASSO logistic regression models were used to predict 30-day readmission. 
Both were tuned via cross-validation, yielding similar performance (accuracy ≈ 91.6%, 
AUC ≈ 0.606) . 

Due to class imbalance (only ~8.4% readmitted), the default 0.5 threshold resulted in 
both models predicting no readmissions, with 0% sensitivity and 100% specificity.
This high accuracy is therefore misleading, as it reflects the majority class rather than true predictive ability. 

The modest AUC indicates limited discrimination. LASSO additionally performed variable 
selection, highlighting key predictors such as discharge status and time in hospital, 
while ridge retained all variables with shrinkage. 

In this context, missing readmissions is more costly than false positives, suggesting 
that a lower threshold or alternative evaluation approach would improve model usefulness.

## Generalized Additive Model

```{r }
# Generalized Additive Model

#Create a train and test set to evaluate the test error rate for each 
# version of the gam

set.seed(1)
gam_model_data <- model_data
gam_model_data$primary_diagnosis_group <- unname(gam_model_data$primary_diagnosis_group)


gam_train_index <- createDataPartition(
  gam_model_data$readmitted_binary,
  p = 0.7,
  list = FALSE
)

gam_train_data <- gam_model_data[gam_train_index, ]
gam_test_data  <- gam_model_data[-gam_train_index, ]

#First, make a gam with no nonlinear terms, trained only on the training data
gam1 <- gam(readmitted_binary ~ ., family=binomial, data=gam_model_data, 
            subset=gam_train_index)

#Then, make a second gam
#Here, we will have a nonlinear term only for time_in_hospital, trained only on
# training data
gam2 <- gam(readmitted_binary ~ s(time_in_hospital,df=4) +
            hba1c_group + gender + age_group_paper + race + 
            admission_source_group + discharge_home + primary_diagnosis_group +
            medical_specialty_group, family=binomial, data=gam_model_data, 
            subset=gam_train_index)

#Calculate test error rates for gam1 and gam2
gam1_probs <- predict(gam1, newdata=gam_test_data, type="response")
gam1_preds <- ifelse(gam1_probs > 0.5, 1, 0)
gam1_test_error <- mean(gam1_preds != gam_test_data$readmitted_binary)
gam1_test_error
# 0.08666747

gam2_probs <- predict(gam2, newdata=gam_test_data, type="response")
gam2_preds <- ifelse(gam2_probs > 0.5, 1, 0)
gam2_test_error <- mean(gam2_preds != gam_test_data$readmitted_binary)
gam2_test_error
# 0.08666747

# This shows that adding a nonlinear effect for time_in_hospital variable does
# not affect the test error rate.
# This means that, in this case, using a GAM with a nonlinear term does not 
# improve the model's prediction beyond logistic regression.


#Creating a third gam with 5 knots on the time_in_hospital nonlinear effect
gam3 <- gam(readmitted_binary ~ s(time_in_hospital,df=5) +
            hba1c_group + gender + age_group_paper + race + 
            admission_source_group + discharge_home + primary_diagnosis_group +
            medical_specialty_group, family=binomial, data=gam_model_data, 
            subset=gam_train_index)


gam3_probs <- predict(gam3, newdata=gam_test_data, type="response")
gam3_preds <- ifelse(gam3_probs > 0.5, 1, 0)
gam3_test_error <- mean(gam3_preds != gam_test_data$readmitted_binary)
gam3_test_error
#  0.08666747

# Overall, these three gams all have the same test error rate, which
# indicates that time_in_hospital likely has an approximately linear effect
# so, giving it a nonlinear effect does not improve the test error rate of
# the model, regardless of how many knots there are.

#plotting the effect of time_in_hospital with and without nonlinear effects
#time <- seq(min(gam_model_data$time_in_hospital), 
 #           max(gam_model_data$time_in_hospital),
#            length.out=200)
plot(gam1 ,se=T,col="green", main="Plot of time_in_hospital effect for Gam1")
plot(gam2 ,se=T,col="red",main="Plot of time_in_hospital effect for Gam2")
plot(gam3 ,se=T,col="lightblue",main="Plot of time_in_hospital effect for Gam3")

#Plot the error rates of Gams with df = 0-10
gam_test_error_vect <- c(rep(0,10))
for (i in 0:10){
  gami <- gam(readmitted_binary ~ s(time_in_hospital,df=i) +
            hba1c_group + gender + age_group_paper + race + 
            admission_source_group + discharge_home + primary_diagnosis_group +
            medical_specialty_group, family=binomial, data=gam_model_data, 
            subset=gam_train_index)
  gami_probs <- predict(gam3, newdata=gam_test_data, type="response")
  gami_preds <- ifelse(gam3_probs > 0.5, 1, 0)
  gami_test_error <- mean(gam3_preds != gam_test_data$readmitted_binary)
  gam_test_error_vect[i] <- gami_test_error
}

table(gam_test_error_vect)
#gam_test_error_vect
#0.0866674725009066 
#                10 

# Gams with df values 1-10 for the time_in_hospital effect all have the 
# same test error rate, which further supports the notion that a 
# nonlinear effect for time_in_hospital is not useful,
# regardless of the number of knots


```

```{r}
### KNN

cluster_vars <- c(
  "hba1c_group", "age_group_paper", "race", "gender",
  "admission_source_group", "discharge_home",
  "primary_diagnosis_group", "medical_specialty_group",
  "time_in_hospital"
)

cat_vars <- setdiff(cluster_vars, "time_in_hospital")

model_data <- diabetic_clean %>%
  drop_na(all_of(cluster_vars)) %>%
  drop_na(readmitted_binary) %>%
  mutate(across(all_of(cat_vars), as.factor))

# one-hot encode categorical variables
encoded <- model.matrix(~ . - 1, data = model_data[, cat_vars])

# scale all columns so no single variable dominates distance calculation
# time_in_hospital would otherwise outweigh binary 0/1 columns
X <- scale(cbind(encoded, time_in_hospital = model_data$time_in_hospital))
y <- factor(model_data$readmitted_binary, levels = c(0, 1))

cat("Readmission rate:", round(mean(model_data$readmitted_binary), 3), "\n")

```

```{r}
# STEP 2: Split into train (70%), validation (15%), test (15%)
#   Train: used to fit KNN
#   Validation: used to pick the best k (k=1 through 20)
#   Test: held out entirely until final evaluation — never seen during tuning
    
set.seed(42)
n <- nrow(X)
train_idx <- sample(n, size = floor(0.70 * n))
remaining  <- setdiff(1:n, train_idx)
val_idx    <- sample(remaining, size = floor(0.50 * length(remaining)))
test_idx   <- setdiff(remaining, val_idx)

X_train <- X[train_idx, ];  y_train <- y[train_idx]
X_val   <- X[val_idx, ];    y_val   <- y[val_idx]
X_test  <- X[test_idx, ];   y_test  <- y[test_idx]

cat("Train:", length(train_idx), "| Validation:", length(val_idx), "| Test:", length(test_idx), "\n")
```

```{r}
# STEP 3: Test k = 1 through 20 on validation set
#   We use AUC rather than accuracy to pick k because readmission is imbalanced
results <- data.frame(k = 1:20, accuracy = NA, sensitivity = NA, specificity = NA, auc = NA)

for (k in 1:20) {
  pred <- knn(X_train, X_val, cl = y_train, k = k, prob = TRUE)
  prob_readmitted <- ifelse(pred == 1, attr(pred, "prob"), 1 - attr(pred, "prob"))
  cm <- confusionMatrix(pred, y_val, positive = "1")
  roc_obj <- roc(as.numeric(y_val) - 1, prob_readmitted, quiet = TRUE)
  results$accuracy[k]    <- round(cm$overall["Accuracy"], 4)
  results$sensitivity[k] <- round(cm$byClass["Sensitivity"], 4)
  results$specificity[k] <- round(cm$byClass["Specificity"], 4)
  results$auc[k]         <- round(auc(roc_obj), 4)
}

best_k <- results$k[which.max(results$auc)]
cat("Best k:", best_k, "| AUC:", max(results$auc), "\n")
```
```{r}
# pick best k by AUC
best_k <- results$k[which.max(results$auc)]
cat("\nBest k:", best_k, "| AUC:", max(results$auc), "\n")

# plot k selection
ggplot(results, aes(x = k)) +
  geom_line(aes(y = auc, color = "AUC")) +
  geom_line(aes(y = sensitivity, color = "Sensitivity")) +
  geom_line(aes(y = specificity, color = "Specificity")) +
  geom_vline(xintercept = best_k, linetype = "dashed", color = "black") +
  annotate("text", x = best_k + 0.5, y = 0.5, label = paste("Best k =", best_k), hjust = 0) +
  scale_color_manual(values = c("AUC" = "steelblue", "Sensitivity" = "tomato", "Specificity" = "seagreen")) +
  labs(
    title = "KNN Validation Performance Across k = 1 to 20",
    subtitle = "Best k selected by highest AUC",
    x = "k (number of neighbors)", y = "Score", color = "Metric"
  ) +
  theme_minimal()
```
```{r}
# STEP 4: Fit final KNN with best k on test set
#   Test set was never used during k selection — this is an unbiased estimate
#   of how the model would perform on new, unseen patients

final_pred <- knn(X_train, X_test, cl = y_train, k = best_k, prob = TRUE)
final_probs <- ifelse(final_pred == 1, attr(final_pred, "prob"), 1 - attr(final_pred, "prob"))
cm_final  <- confusionMatrix(final_pred, y_test, positive = "1")
roc_final <- roc(as.numeric(y_test) - 1, final_probs, quiet = TRUE)
cat("FINAL MODEL PERFORMANCE (Test Set) \n")
cat("k =", best_k, "\n")
cat("Accuracy:    ", round(cm_final$overall["Accuracy"], 4), "\n")
cat("Sensitivity: ", round(cm_final$byClass["Sensitivity"], 4),
    " ← % of readmitted patients correctly identified\n")
cat("Specificity: ", round(cm_final$byClass["Specificity"], 4),
    " - % of non-readmitted patients correctly cleared\n")
cat("AUC:         ", round(auc(roc_final), 4),
    " - overall discrimination (0.5 = random, 1.0 = perfect)\n")
# confusion matrix heatmap
conf_df <- as.data.frame(cm_final$table)
colnames(conf_df) <- c("Predicted", "Actual", "Count")
ggplot(conf_df, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), size = 7, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  scale_x_discrete(labels = c("0" = "Not Readmitted", "1" = "Readmitted")) +
  scale_y_discrete(labels = c("0" = "Not Readmitted", "1" = "Readmitted")) +
  labs(
    title = "Confusion Matrix — KNN Readmission Prediction",
    subtitle = paste0("k = ", best_k,
                      " | Accuracy: ", round(cm_final$overall["Accuracy"] * 100, 1), "%",
                      " | AUC: ", round(auc(roc_final), 3))
  ) +
  theme_minimal()
# ROC curve
roc_df <- data.frame(
  specificity = rev(roc_final$specificities),
  sensitivity = rev(roc_final$sensitivities)
)
ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_abline(linetype = "dashed", color = "gray50") +
  annotate("text", x = 0.7, y = 0.2,
           label = paste0("AUC = ", round(auc(roc_final), 3)),
           size = 5, color = "steelblue") +
  labs(
    title = "ROC Curve — KNN Readmission Prediction",
    subtitle = "Dashed line = random guessing (AUC = 0.5)",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal()
```